Dynastic Extinctions and Urban Economic Development: \\Evidence from the Holy Roman Empire\footnote{I thank Davide Cantoni, Matthias Weigand, and my RA colleagues at the Chair of Economic History for helpful comments and feedback.}

### 2023-06-09 paper/paper.tex
#22 finish introduction

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Their model explains why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439).
However, it focuses more on these aggregate patterns than on the traits that cause states to succeed.
\footnote{The basic idea of Darwinian selection has previously been borrowed by economists, though - for example \cite{galor2002} do the literal version, where humans are subject to selection and accumulate favourable traits over the course of history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that the rich out-bred the poor and passed favourable traits on to their offspring. in Europe but not elsewhere.}

@@

My story is not about population but institutions (although it is compatible with Malthusianism). A longstanding tradition in economics emphasises the role of institutional factors in long-run growth (e.g. \cite{north1970}, \cite{delong1993}, \cite{ajr2001}). From this vast literature one theme is especially relevant here: the institution of fiscal capacity, or the state's ability to collect taxes. \cite{tilly1985} argues that the main reason for states to collect taxes is to be able to wage war. In other words, fiscal capacity and state power are tightly linked --- if conflict selects for state power, then it also selects for fiscal capacity. \cite{cantoni2023} show that drove state consolidation and played a role in growth. \cite{dincecco2012}.

@@

The closest relative to this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict.

@@

The rest of the paper proceeds as follows.

\subsection*{Other literature}

\cite{levine2013, levine2021} provide the theoretical foundation for conflict-driven selection, as well as compelling evidence for it. 
They argue that social evolution is driven by conflict. Their model finds that "evolution selects societies with high state power - the ability to prevail in conflict over rivals." (\citealp{levine2021}, p. 436).Their answer to Why Europe and not China or India: "Ghengis Khan, the cannon, and the English. Ghengis Khan accomplished two things: he brought cannons to Europe, and he depopulated Mongolia. In China the removal of the outside threat of the Mongols led - as the theory asserts - to an extractive hegemony. In Europe the advent of the cannon together with the preying of the English on the French and the Spanish resulted - as the theory asserts - in an inclusive balance of power." (p. 439). strong outsiders but primitive siege technology [cheap defensibility of cities] in India led to extractive balance of power.

How they define state power (pp. 439-440): "the potential ability of a social organisation to resist and influence other societies. This depends on institutional characteristics of the society, including stability of the government and the ability to collect taxes and conscript soldiers, which depend in turn on the law-abidingness of citizens, the efficiency of the courts, and the overall economic strength of the society." They *assume* in their model (p.443) that state power is directly proportional to free resources (resources in excess of subsistence).

I do not study their explanation for the European equilibrium that we observe in history; rather, I take a step back by taking the state system and the process of conflict-driven evolution as given. What interests me are the consequences of this process for economic growth.
Is it, as \cite{schoenholzer2022} argue explicitly, beneficial creative destruction - does state power equal state quality? Or are powerful states detrimental after all - because they engage in more extraction? \cite{tilly1985}

Why we would expect the hypothesis to be true: state power, defined as the ability to win conflicts with rivals, consists of factors like fiscal capacity, efficient courts, law-abiding citizens (\citealp{levine2021}, pp.439-449) - institutional factors that are also widely held to be beneficial for growth (see the vast literature on institutions). Add some institutions lit here.

My central identifying assumption: conquering a city implies more state power than inheriting it.

\cite{schoenholzer2022} use a dataset with a bigger geographical scope but lower resolution: they use urban population from \cite{bairoch1988} as a proxy for development, which is only available in intervals of 50 or 100 years, and a series of political maps.

Key motivation papers: \cite{levine2013, levine2021}, \cite{diamond1997}, \cite{schoenholzer2022}, \cite{cervellati2022}, \cite{voigtlnder2013}, \cite{tilly1985}

The idea that today's states are just the most successful protection rackets of the past already shows up in \cite{tilly1985}.

While frequent wars brought devastation over the people and destruction to property, \cite{malthus1798} and others after him have argued, frequent wars may have kept population in check, leading to higher land-labour ratios and therefore a higher level of income per capita. \cite{voigtlnder2013} argue that the relative wealth of Europe after 1350 can be explained by a feedback loop centered on this Malthusian mechanism: after the huge loss of life caused by the Black Death, which served as a catalyst, taxable incomes rose, which made rulers richer, which allowed them to afford more wars, which in turn kept population in check.

>>>>

I build my paper on the theory of conflict driven evolution, outlined above, developed by \cite{levine2013, levine2021, levine2022}. Its main contribution is to explain why hegemonies arose in East Asia while balances of power could prevail in Europe: unlike China, the major powers of continental Europe were frequently preyed upon by strong outsiders --- the Vikings, the Swedes, and the English --- which prevented any one state from growing too strong (\citealp{levine2021}, p. 439). It is worth highlighting how the authors model war between states: you win with state power, and state power is simply free resources, i.e. resources above subsistence. If the most successful states are the ones with the biggest surplus, selection should have made Europe increasingly prosperous.\footnote{The basic idea of Darwinian selection has previously been borrowed by economists - for example \cite{galor2002} do the literal version, where humans are subject to selection and accumulate favourable traits over the course of history. More infamously, \cite{clark2007} employs it in an attempt to explain why the Industrial Revolution took place in Europe, arguing that the rich out-bred the poor and passed favourable traits on to their offspring (in Europe but not elsewhere for some reason.)}

@@

My story is not about population but institutions (although it is compatible with Malthusianism). A longstanding tradition in economics emphasises the role of institutional factors in long-run growth (e.g. \cite{north1970}, \cite{delong1993}, \cite{ajr2001}). From this vast literature one theme is especially relevant here: the institution of fiscal capacity, or the state's ability to collect taxes. \cite{tilly1985} argues that the main reason for states to collect taxes is to be able to wage war. In other words, fiscal capacity and state power are tightly linked --- so if conflict selects for state power, then it also selects for fiscal capacity. \cite{cantoni2023} show that conflict drove state consolidation. \cite{dincecco2012} show that fiscal capacity had a positive effect on growth.

@@

The closest relative to this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict. They use a dataset with a bigger geographical scope (all of Europe) but lower resolution: they use urban population from \cite{bairoch1988} as a proxy for development, which is only available in intervals of 50 or 100 years, and a series of political maps.

@@

The rest of the paper proceeds as follows. Section 2 presents some background information. Section 3 describes the data used in the analysis. Section 4 details the empirical approach. Section 5 presents results. Section 6 discusses implications and robustness of the results. Section 7 concludes.
